New modal operators for constructive mathematics

### 2023-09-11 slides-herrsching2023.tex
Slides for first talk finished

  Then for every parameter value~$t$, classically there is

@@

  \begin{columns}[c]
    \begin{column}{0.01\textwidth}
      \includegraphics[height=2.4em]{question-mark}
    \end{column}
    \begin{column}{0.9\textwidth}
      \mbox{Can locally the functions~$\lambda_i$ be chosen to be continuous?
      \only<2->{\hil{Yes.}}} \\
      How about the~$v_i$? \only<2->{\hil{No.}}
    \end{column}
  \end{columns}
\end{frame}

@@

@@

  \begin{columns}
    \begin{column}{0.45\textwidth}
      \begin{block}{1. Explore foundational possibility}
        There are forcing extensions with~\textsc{ch}, $\neg$\textsc{ch},
        \textsc{lem}, $\neg$\textsc{lem}, \ldots
      \end{block}
    \end{column}
    \pause

@@

  \textbf{Def.} For a predicate~$P$ on finite lists over a set~$X$, inductively
  define:
  \[
    \infer{P \,|\, \sigma}{P\sigma}
    \qquad
    \infer{P \,|\, \sigma}{\forall(x \in X)\_\ P \,|\, \sigma x}
  \]

  \textbf{Def.} A preorder is \hil{well} iff~$\mathsf{Good} \,|\, [\,]$, where
  $\mathsf{Good}\,[\sigma_1,\ldots,\sigma_n] \defeqv (\exists(i < j)\_ \sigma_i \leq \sigma_j)$.
\end{frame}}

@@

>>>>

  Then for every parameter value~$t \in \Omega$, classically there is

@@

  \begin{columns}[c]
    \begin{column}{0.01\textwidth}
      \includegraphics[height=2.4em]{question-mark}
    \end{column}
    \begin{column}{0.9\textwidth}
      \mbox{Can locally the functions~$\lambda_i$ be chosen to be continuous?
      \only<2->{\hil{Yes.}}} \\
      How about the~$v_i$? \only<2->{\hil{No}\only<3->{, but \hil{yes} on a dense
      open subset of~$\Omega$.}}
    \end{column}
  \end{columns}
\end{frame}

@@

\section{What can forcing do for you?}

@@

  \begin{columns}
    \begin{column}{0.45\textwidth}
      \begin{block}{1. Explore foundational possibility}
        There are forcing extensions with \\
        \textsc{ch}, $\neg$\textsc{ch}, \textsc{lem}, $\neg$\textsc{lem}, \ldots
      \end{block}
    \end{column}
    \pause

@@

  \textbf{Def.} A preorder is \hil{well} iff any of the following equivalent
  conditions hold:
  \begin{enumerate}
    \item The \hil{generic sequence} $\NN \to X$ is good.
    \pause
    \item Every sequence $\NN \to X$ \hil{in every forcing extension} is good.
    \pause
    \item There is a \hil{well-founded tree} witnessing universal goodness.
  \end{enumerate}

%  $\mathsf{Good}\,[\sigma_1,\ldots,\sigma_n] \defeqv (\exists(i < j)\_ \sigma_i \leq \sigma_j)$.
%  \textbf{Def.} For a predicate~$P$ on finite lists over a set~$X$, inductively
%  define:
%  \[
%    \infer{P \,|\, \sigma}{P\sigma}
%    \qquad
%    \infer{P \,|\, \sigma}{\forall(x \in X)\_\ P \,|\, \sigma x}
%  \]
%
%  \textbf{Def.} A preorder is \hil{well} iff~$\mathsf{Good} \,|\, [\,]$, where
%  $\mathsf{Good}\,[\sigma_1,\ldots,\sigma_n] \defeqv (\exists(i < j)\_ \sigma_i \leq \sigma_j)$.
\end{frame}}

\section{Basics of forcing}

\begin{frame}{Forcing notions}
  \small
  \textbf{Def.} A \hil{forcing notion} consists of
  a preorder~$L$ of \hil{forcing conditions}, and
  for every~$\sigma \in L$, a set~$\Cov(\sigma) \subseteq
  P({\downarrow}\sigma)$ of \hil{coverings} of~$\sigma$
  such that: If~$\tau \preccurlyeq \sigma$ and~$R \in \Cov(\sigma)$, there
  should be a covering~$S \in \Cov(\tau)$ such that~$S \subseteq
  {\downarrow}R$.
  \bigskip

  {\centering\footnotesize\begin{tabular}{llll}
    \toprule
    & preorder~$L$ & coverings of an element~$\sigma \in L$ & filters of~$L$ \\
    \midrule
    \normalnumber{1} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$ & maps~$\NN \to X$ \\
    \normalnumber{2} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$,\ \ $\{ \sigma\tau \,|\, \tau \in X^*, a \in \sigma\tau \}$ for each~$a \in X$ & surjections~$\NN \twoheadrightarrow X$ \\
    \normalnumber{3} & f.g. ideals & --- & ideals \\
    \normalnumber{4} & f.g. ideals & $\{ \sigma+(a), \sigma+(b) \}$ for each~$ab \in \sigma$,\ \ $\{\}$ if~$1 \in \sigma$ & prime ideals \\
    \normalnumber{5} & opens & $\mathcal{U}$ such that~$\sigma = \bigcup \mathcal{U}$ & points \\
    \normalnumber{6} & $\{\star\}$ & $\{ \star \,|\, \varphi \} \cup \{ \star \,|\, \neg\varphi \}$ &
    witnesses of~\textsc{lem}
    \\
    \bottomrule
  \end{tabular}\par}
  \bigskip

  \textbf{Def.} A \emph{filter} of a forcing notion~$(L,\mathrm{Cov})$
  is a subset~$F \subseteq L$ such that
  \vspace*{-0.4em}
  \begin{enumerate}
    \scriptsize
    \item $F$ is upward-closed: if~$\tau \preccurlyeq \sigma$ and if~$\tau \in F$, then~$\sigma \in F$; \\[-3.0em]
    \item $F$ is downward-directed: $F$ is inhabited, and if~$\alpha,\beta \in F$,
    then there is a common refinement~$\sigma \preccurlyeq \alpha,\beta$ such
    that~$\sigma \in F$; and \\[-2.0em]
    \item $F$ splits the covering system: if~$\sigma \in F$ and~$R \in
    \Cov(\sigma)$, then~$\tau \in F$ for some~$\tau \in R$.
  \end{enumerate}

\end{frame}

@@

Forschungsfrage: Klassische Basen sind ja schon ziemlich maximal in dem Sinn,
als dass etwa der Raum der Funktionen ℕ → ℕ genügend Punkte hat. Wie wäre es,
in einer Welt zu arbeiten, in der auch der Raum der Surjektionen ℕ → X genügend
Punkte hat?
